# Identifying DNA fragment that increases 

## Manipulating data tables

We will be using the tidyverse in this lab. So let's start by loading it.

```{r}
library(tidyverse)
```

Your lab conducted in experiment in which four different fragments of chromosome 21 were integrated into mice. The four parts are denoted with _141G6_, _152F7_, _230E8_ and _285E6_. The mice were bred resulting in dozens of  transgenic mice. The DNA fragment is not always inherted so some mice have the extra copy and others don't. We are interested in determining if any of these fragments result in an increase in weight, a characteristic associated with trisomic mice.

The data can be loaded like this.

```{r}
load("rdas/mouse.rda")
```

Which loads the `dat` object into your environment:

```{r}
dat
```

The columns included in this table are the following:

  - _DNA_: Fragment of chromosome 21 integrated in parent mouse (1=141G6; 2=152F7; 3=230E8; 4=285E6).
  - _line_: Family line.
  - _tg_ - Whether the mouse contains the extra DNA (1) or not (0).
  - _sex_:  Sex of mouse (1=male; 0=female).
  - _age_: Age of mouse (in days) at time of weighing.
  - _weight_: Weight of mouse in grams, to the nearest tenth of a gram.
  - _bp_: Blood pressure of the mosue.
  - _cage_: Number of the cage in which the mouse lived
  

Let's start by comparing the weights of the no trisomic mice to the weights of mice with the other four fragmets. Determine which columns tells us the fragment of the mouse and determine the number of mouse in each group? Hint: use the _count_ function.

```{r}
dat %>% count(DNA)
```

Note that the names are 1, 2, 3, 4. Let's change these to the actual names
1=141G6; 2=152F7; 3=230E8; 4=285E6. Create a new columna called `fragrment` with the actual fragment nemas. Hint: Use the `recode` funtion. 


```{r}
dat <- mutate(dat, fragment = recode(DNA, 
                                "1"="141G6", 
                                "2"="152F7", 
                                "3"="230E8", 
                                "4"="285E6"))
```

Note that all the mice in our table have one of these names. However, we know that not all mouse have the fragments. Remember that not all inhereted the extra copy. Use `filter` and `count` to see how many mice in the `141G6` group have the extra copy.

```{r}
filter(dat, fragment == "141G6") %>% count(tg)
```

Now change the `fragment` column so that the mice that do not have the extra copy, have are called  `No trisomy`. Hint: use the `ifelse` function.

```{r}
dat <- dat %>% mutate(fragment = ifelse(tg == 0, "No trisomy", fragment)) 
```

Before we continue let's learn about the `n()` function. Note that we can perform the same as the `count()` function using `group_by()` and `n()`

```{r}
dat %>% group_by(DNA) %>% summarize(freq = n())
```

Now compute the average and standard error in each of the four groups and the control. Hint: Use `group_by` and `summarize`. 

```{r}
dat %>% group_by(fragment) %>% 
  summarize(average = mean(weight), se = sd(weight)/sqrt(n()))
```


Bonus: Is the above difference statisticall signficant at the 0.05 level?
```{r}
lm(weight ~ tg, data = dat) %>% summary() %>% .$coef
```

This is what the dynamite plot would look like:

```{r, echo=FALSE}
dat %>% group_by(fragment) %>% 
  summarize(average = mean(weight), se=sd(weight)/sqrt(n())) %>%
  ggplot(aes(fragment, average)) + 
  geom_errorbar(aes(ymin = average - se, ymax = average+se), width = 0.25)+
  geom_bar(stat = "identity", width=0.5, fill=4, col = 1) +
  xlab("") + ylab("Outcome")
```

Does this make sense? In the next section we demonstrate how data exploration allows us to detect a problem. We also learn how to make publication ready plots.

## Data visualization


### Data exploration to identify outliers

A problem with the summary statistics and the barplot above is that we actually do not see the data. Use the _geom_point_ function to show all the points. Do you see a problem? What is it?

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_point()
```

We know that 1000 gram mice do not exist. In fact 100 grams is already a huge mouse. Use filter to show the data from the mice weighing more than 100 grams. 

```{r}
filter(dat, weight > 100) 
```

What are the weights?

```{r}
filter(dat, weight > 100) %>% pull(weight)
```

An unfortunate common practice is to use the number 999 to denote missing data. The recommended practice is to acutaly type NA. Use the fitler function remove all the rows with these missing values, then remake the figure and recompute the averages and standard errors.

```{r}
dat <- filter(dat, weight != 999)

dat %>% ggplot(aes(fragment, weight)) + geom_point()

dat %>% group_by(fragment) %>% 
  summarize(average = mean(weight), se = sd(weight)/sqrt(n()))
```

This makes much more sense. 

## Data exploration to help answer scientific question

From the table above we now see that 
Make a boxplot
```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot()
```

relevel fragment to have control first

```{r}
dat <- mutate(dat, fragment = factor(fragment)) %>% mutate(fragment = relevel(fragment, ref = "No trisomy"))
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot()
```

add points

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot() + 
  geom_point()
```

add jitter

```{r}
dat %>% ggplot(aes(fragment, weight)) + geom_boxplot() + 
  geom_jitter()
```

smaller spread, alpha-blend 

```{r}
dat %>% ggplot(aes(fragment, weight)) +
  geom_boxplot() + 
  geom_jitter(width = 0.1, alpha = 0.5)
```

a unexpected result is revealed. bimodal

ridge plots


histograms
```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_histogram()
```

avoid warning and make it nice

```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_histogram(binwidth = 1, color = "black")
```

density
```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_density()
```

make it nice

```{r}
filter(dat, fragment == "No trisomy") %>%
  ggplot(aes(weight)) +
  geom_density(fill = 1, alpha = 0.5) + 
  xlim(c(19,45))
```

ridge plots

```{r}
library(ggridges)
dat %>% 
  ggplot(aes(weight, fragment)) +
  geom_density_ridges() 
```
Avoid overal 

```{r}
library(ggridges)
dat %>% 
  ggplot(aes(weight, fragment)) +
  geom_density_ridges(scale = 0.9) 
```

sex?

```{r}
dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5)
```


sex is a number! let's change it to charcater or factor.

```{r}
dat <- mutate(dat, mutate(sex = ifelse(sex == 1, "Male", "Female"))

)
dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5)
```

Facet

```{r}
dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  facet_grid(.~sex)
```

Themes

```{r}
dat %>% ggplot(aes(fragment, weight, color = sex)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  facet_grid(.~sex) + 
  theme_bw()
```

Cage effect?
Are males and females caged together?

```{r}
with(dat, table(cage, sex))
```

Is cage counfounded?

```{r}
dat %>% 
  filter(sex == "Female") %>%
  mutate(cage = factor(cage)) %>%
  ggplot(aes(cage, weight, color = fragment)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_bw()
```

Order by meaningful value


```{r}
dat %>% 
  filter(sex == "Female") %>%
  mutate(cage = factor(cage)) %>%
  mutate(cage = reorder(cage, weight, median)) %>%
  ggplot(aes(cage, weight, color = fragment)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_bw()
```

```{r}
dat %>% 
  filter(sex == "Male") %>%
  mutate(cage = factor(cage)) %>%
  mutate(cage = reorder(cage, weight, median)) %>%
  ggplot(aes(cage, weight, color = fragment)) +
  geom_boxplot(width = 0.5, color = "grey") + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_bw()
```


```{r}
filter(dat, sex == "Female") %>% 
  lm(weight ~ fragment , data = .) %>% summary()
filter(dat, sex == "Female") %>% 
  lm(weight ~ fragment , data = .) %>% aov() %>% summary()

```

