---
title: "Vizualization with ggplot2"
author: "Rafael A. Irizarry"
date: "3/1/2019"
output:
  ioslides_presentation:
    fig_caption: no
    fig_height: 4
    fig_width: 7
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(dslabs)
library(ggthemes)
library(ggrepel)

data(murders)
```

## Introduction

- There are also other packages for creating graphics such as `grid` and `lattice`, but `ggplot2` provides intuitive and powerful tools to create plots. 
- `ggplot2` breaks plots into components in a way that permits beginners to create relatively complex and aesthetically pleasing plots using syntax that is intuitive and comparatively easy to remember.
- It provides intuitive tools to create hundreds of different plots by using the [grammar of graphics](https://www.springer.com/us/book/9780387245447) (the gg in `ggplot2`)
- Make use of the `ggplot2` [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)!

## Introduction
- Motivation: You will know how to create this plot

```{r ggplot-example-plot, fig.cap = "Murder totals versus population size for US states.", echo=FALSE}
r <- murders %>% 
  summarize(pop=sum(population), tot=sum(total)) %>% 
  mutate(rate = tot/pop*10^6) %>% .$rate

murders %>% ggplot(aes(x = population/10^6, y = total, label = abb)) +  
  geom_abline(intercept = log10(r), lty=2, col="darkgrey") +
  geom_point(aes(color=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name="Region") +
  theme_economist()
```

## The Components of a Graph

The main three components of a graph are:

- **Data**: The US murders data table is being summarized. We refer to this as the **data** component.
- **Geometry**: The plot above is a scatterplot. This is referred to as the geometry component.
- **Aesthetic mapping**: The plot uses several visual cues to represent the information provided by the dataset. The two most important cues in this plot are the point positions in the x-axis and y-axis, which represent population size and the total number of murders respectively. Each point represents a different observation which implies we need to *map* the data to the visual cue. Color is another visual cue that represents region and also needs a map. We refer to this as the **aesthetic mappings** component. How we define the mapping depends on what **geometry** we are using.

## `ggplot2` objects: A Blank Slate
- Start by loading the `dslabs` package.

```{r}
ggplot(data = murders)
```

## `ggplot2` objects: A Blank Slate

- We can also use the pipe operator:

```{r}
murders %>% ggplot()
```

## `ggplot2` objects: A Blank Slate

- Why is it "blank"?

## `ggplot2` objects: A Blank Slate

- Why is it "blank"?

The code renders a blank slate since no geometry has been defined

## `ggplot2` objects: A Blank Slate

- Why is it "blank"?

The code renders a blank slate since no geometry has been defined

- We can define a `ggplot2` object like this:
```{r}
p <- ggplot(data = murders)
class(p)
```

- Finally, we can see the object with:

```{r eval=FALSE}
print(p)
p 
```

## Geometries

- We create graphs by adding *layers* 
- Layes can define geometries, summary statistics, styles, among other things
- We use `+` to add layers
- In general, a line of code looks like this:


>> DATA %>% `ggplot()` + LAYER 1 + LAYER 2 + ... + LAYER N

- Suppose we want to create a scatter plot. Taking a look at the [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) we see that we need to use `geom_point` 

- Note that geometry function names follow this pattern: `geom` and the name of the geometry connected by an underscore.

## Aesthetic Mappings 

- The function `aes` connects data with what we see on the graph. 
- We refer to this as the **aesthetic mappings**.
- This example produces a scatterplot of total murders versus population in millions:
```{r eval=FALSE}
murders %>% ggplot() + 
  geom_point(aes(x = population/10^6, y = total))
```

- We can drop the `x =` and `y =` if we wanted to since these are the first and second expected arguments

- We can also add a layer to the `p` object that was defined before as `p <- ggplot(data = murders)`

## Aesthetic Mappings

```{r}
p + geom_point(aes(population/10^6, total))
```


## Layers

- Recall the figure we are aiming to recreate:

```{r echo=FALSE}
r <- murders %>% 
  summarize(pop=sum(population), tot=sum(total)) %>% 
  mutate(rate = tot/pop*10^6) %>% .$rate

murders %>% ggplot(aes(x = population/10^6, y = total, label = abb)) +  
  geom_abline(intercept = log10(r), lty=2, col="darkgrey") +
  geom_point(aes(color=region), size = 3) +
  geom_text_repel() + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name="Region") +
  theme_economist()
```

- A second layer we must add is the label to each point to identify the state.

## Layers
- The `geom_label` and `geom_text` functions allow us to add text to plots
```{r}
p + geom_point(aes(population/10^6, total)) +
  geom_text(aes(population/10^6, total, label = abb))
```
- We have successfully added a second layer to the plot

## Tinkering with Arguments
- Each geometry has arguments other than `aes` and `data`. For example, we can change the size of points:
```{r}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb))
```

## Tinkering with Arguments
- Its hard to see the labels. We can use the `nudge_x` argument to move the text slightly:
```{r}
p + geom_point(aes(population/10^6, total), size = 3) +
  geom_text(aes(population/10^6, total, label = abb), nudge_x = 1)
```

## Global vs Local Aesthethic Mappings
- Notice that we defined the same aesthetic mapping twice:

```{r eval = FALSE}
p + geom_point(aes(population/10^6, total)) +
  geom_text(aes(population/10^6, total, label = abb))
```

- If we define a mapping in ggplot, then all the geometries that are added as layers will default to this mapping. We redefine `p`:

```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb))
```

- Then we can simply write the following code:

```{r eval = FALSE}
p + geom_point(size = 3) + 
  geom_text(nudge_x = 1.5)
```

## Global vs Local Aesthethic Mappings
- Here is the resulting plot:

```{r echo = FALSE}
p + geom_point(size = 3) + 
  geom_text(nudge_x = 1.5)
```

- Why are the `size` and `nudge_x` arguments not defined withing the `ggplot` function?

## Global vs Local Aesthethic Mappings
- We can override the global mapping by defining a new mapping within each layer. Here is an example:

```{r}
p + geom_point(size = 3) +  
  geom_text(aes(x = 10, y = 800, label = "Hello there!"))
```

## Scales
- Recall that the desired *scales* are in log-scales. Since this is not default, we need to change it. We can do this with `scale_x_continuous` or with `scale_x_log10`.
- Here is the code:

```{r eval= FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10") 
```

```{r eval= FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() 
```

- `nudge_x` is made smaller due to the log-scale

## Scales

- Here is the resulting plot:

```{r}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() 
```

## Labels & Titles

- The [cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) reveals how to add labels and titles.
- Here is the code

```{r eval = FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

- We are almost done!

## Labels & Titles

- Here is the resulting plot

```{r echo = FALSE}
p + geom_point(size = 3) +  
  geom_text(nudge_x = 0.05) + 
  scale_x_log10() +
  scale_y_log10() +
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")
```

- All we have left to do is add color, a legend and optional changes to the style.

## Categories as Colors

- We can change the color of the data points using the `col` argument in the `geom_point` function. 
- For simplicity, let us redefined `p` as:

```{r eval = FALSE}
p <-  murders %>% ggplot(aes(population/10^6, total, label = abb)) +   
        geom_text(nudge_x = 0.05) + 
        scale_x_log10() +
        scale_y_log10() +
        xlab("Populations in millions (log scale)") + 
        ylab("Total number of murders (log scale)") +
        ggtitle("US Gun Murders in 2010")
```

- For example, here is the code to make the data points blue:

```{r eval = FALSE}
p + geom_point(size = 3, color ="blue")
```

## Categories as Colors










