
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

# Data Wrangling: How to Clean your Data

We will be using the `tidyverse` in this lab. So let's start by loading it.

```{r}
library(tidyverse)
```

Suppose your lab conducts an experiment in which four different fragments of chromosome 21 were integrated into mice. The four parts are denoted with _141G6_, _152F7_, _230E8_ and _285E6_. The mice were bred resulting in dozens of  transgenic mice. The DNA fragment is not always inherted so some mice have the extra copy and others don't. Furthermore, suppose the data was stored as two datasets: **weight.csv** and **blood-pressure.csv**. The goal of this lecture is to clean and merge the datasets in order to end up with this:

```{r echo = FALSE}
load("rdas/mouse.rda")
as_tibble(dat) %>% head() %>% knitr::kable()
```

The columns included in this table are the following:

  - _DNA_: Fragment of chromosome 21 integrated in parent mouse (1=141G6; 2=152F7; 3=230E8; 4=285E6).
  - _line_: Family line.
  - _tg_ - Whether the mouse contains the extra DNA (1) or not (0).
  - _sex_:  Sex of mouse (1=male; 0=female).
  - _age_: Age of mouse (in days) at time of weighing.
  - _weight_: Weight of mouse in grams, to the nearest tenth of a gram.
  - _bp_: Blood pressure of the mosue.
  - _cage_: Number of the cage in which the mouse lived

Let's start by loading the data and looking at it

```{r message=FALSE, warning=FALSE}
library(readxl)
bp     <- read_excel("data/mouse-raw-data.xlsx", sheet=1)
weight <- read_excel("data/mouse-raw-data.xlsx", sheet=2) 
```


Now we can look at the data. Here is a sample of **blood-pressure.csv**:

```{r echo=FALSE}
as_tibble(bp) %>% head() %>% knitr::kable()
```

Each row in this dataset correspond to a mouse. The first three columns describe the family line of the mouse and the last columns corresponds to a blood pressure measurement. Here is a sample of **weight.csv**:

```{r echo=FALSE}
as_tibble(weight[,1:6]) %>% head() %>% knitr::kable()
```

The column names in this table correspond to the characteristics of mice at each DNA fragment. Here we show a sample of observations that have the _141G6_ fragment. Try looking at **weight.csv** for yourself and notice the other columns. We need to clean a lot of data! 

## Wrangling BP dataset

Let's start with **blood-pressure.csv**. The first thing we want to do is to create a new variable that is the combination of the first three columns of the table. We will call this new variable `line`

```{r}
bp <- bp %>%
      mutate(line = paste(line.1, line.2, line.3, sep = "-"))
```

```{r echo=FALSE}
bp %>% head() %>% knitr::kable()
```

Notice that the `line` data in **weight.csv** has a # symbol at the beginning. Let's add that:

```{r}
bp <- bp %>%
      mutate(line = paste("#", line, sep = ""))
```

```{r echo=FALSE}
bp %>% head() %>% knitr::kable()
```

**QUESTION**: Why would we want these to be the same? Recall that the end goal is to merge the two datasets. Furthermore, notice that the only variable in common between the two datasets is `line`. Therefore, we will use `line` to merge the two datasets.

Now, remove the columns `line.1`, `line.2`, and `line.3` from the data. **HINT**: Use the function `select`.

```{r answer 1}
bp <- bp %>% select(line, bp)
```

```{r echo = FALSE}
bp %>% head() %>% knitr::kable()
```

## Wrangling Weight dataset

Let's now move onto **weight.csv**. Recall that the names in this table correspond to the characteristics of mice at each DNA fragment. You can see this by printing the column names:

```{r}
colnames(weight)
```

To start easy, let's consider only the data regarding the _141G6_ fragment. How do you do this?

```{r answer 2, eval = FALSE}
weight %>% 
  select(contains("141G6"))
```

```{r, echo = FALSE}
weight %>% 
  select(contains("141G6")) %>%
  head() %>%
  knitr::kable()
```

Any ideas on how to proceed? Do we need to do much more? 

No! Notice that all the data regarding this DNA fragment is in that table. All we need to do know is change the name of the variables and create a new variable `DNA`. Try doing this by yourself. **Hint**: Use `rename` and `mutate`:

```{r answer 3, eval=FALSE}
weight %>% 
  select(contains("141G6")) %>%
  rename(tg = "141G6.tg",
         sex    = "141G6.sex",
         age    = "141G6.age",
         weight = "141G6.weight",
         cage   = "141G6.cage",
         line   = "141G6.line") %>%
  mutate(DNA = "141G6") 
```

```{r, echo=FALSE}
weight %>% 
  select(contains("141G6")) %>%
  rename(tg = "141G6.tg",
         sex    = "141G6.sex",
         age    = "141G6.age",
         weight = "141G6.weight",
         cage   = "141G6.cage",
         line   = "141G6.line") %>%
  mutate(DNA = "141G6") %>%
  head() %>%
  knitr::kable()
```

Now, repeat this procedure for all DNA fragments. **Hint**: Use loops

```{r}
fragments <- c("141G6", "152F7", "230E8", "285E6")

res <- lapply(seq_along(fragments), function(x){

  tmp <- weight %>%
            select(contains(fragments[x])) %>%
            na.omit() %>%
            rename(tg     = paste(fragments[x], ".tg", sep = ""),
                   sex    = paste(fragments[x], ".sex", sep = ""),
                   age    = paste(fragments[x], ".age", sep = ""),
                   weight = paste(fragments[x], ".weight", sep = ""),
                   cage   = paste(fragments[x], ".cage", sep = ""),
                   line   = paste(fragments[x], ".line", sep = "")) %>%
            mutate(DNA = fragments[x])
})

weight <- do.call(rbind, res)
```

```{r, echo=FALSE}
weight %>%
  head() %>%
  knitr::kable()
```

We are almost done. Recall from above that `DNA` is encoded as: 
 
 - 1 = 141G6
 - 2 = 152F7
 - 3 = 230E8
 - 4 = 285E6

We can use the function `recode` to achieve this:

```{r}
weight <- weight %>%
            mutate(DNA = recode(DNA, "141G6"="1", "152F7"="2", "230E8"="3", "285E6"="4"))
```

```{r echo = FALSE}
weight %>%
  head() %>%
  knitr::kable()
```

Finally, the last step is to merge the two datasets. Recall that we want to use the `line` variable to merge the tables. Give it a try, call the new table `dat`. **Hint**: Use the `left_join` function 

```{r}
dat <- weight %>%
          left_join(bp, by = "line") %>%
          select(DNA, line, tg, sex, age, weight, bp, cage)
```

```{r echo=FALSE}
dat %>%
  head() %>%
  knitr::kable()
```





# Exercises 

1. Check the variable types of each column. Does it make sense? Should we change the variable type of any of the columns?

We can assess this by either printing the data frame or using the `class` function

```{r}
sapply(dat, class)
```

Note that the varible `tg` is `numeric` but it only takes on either 0 or 1. We can convert this to either a character or a factor:

```{r}
dat <- dat %>%
          mutate(tg = as.character(tg))
```


2. Compute the number of mice in each DNA fragment group. Are the number of observations the same? **Hint**: Use `group_by` and `summarize`

```{r, eval = FALSE}
dat %>%
  group_by(DNA) %>%
  summarize(frequency = n()) 
```

```{r, echo = FALSE}
dat %>%
  group_by(DNA) %>%
  summarize(frequency = n()) %>%
  knitr::kable()
```

The number of observations are **not** equal

3. Above we used `weight %>% left_join(bp, by = "line")` to merge the two tables. What happens when you run `bp %>% left_join(weight, by = "line")`? What about `weight %>% right_join(bp, by = "line")`? Do we end up with the same results? **Hint**: What are the row numbers of the tables? Use `anyNA` and `na.omit`


Let's take a look at the number of rows of each table:

```{r}
tmp <- c(nrow(weight), nrow(bp))
names(tmp) <- c("weight", "bp")
tmp
```

The number of rows are different. This implies that when mergin, we have to be aware of missing values. Notice that missing values occur when you run either `bp %>% left_join(weight, by = "line")` or `weight %>% right_join(bp, by = "line")`:

```{r}
weight %>% 
  right_join(bp, by = "line") %>%
  anyNA()

bp %>% 
  left_join(weight, by = "line") %>%
  anyNA()
```

Actually `weight %>% right_join(bp, by = "line")` and `bp %>% left_join(weight, by = "line")` returns the same result. We fix this by removing `NA` with `na.omit`:

```{r eval=FALSE}
weight %>% 
  right_join(bp, by = "line") %>%
  anyNA()

bp %>% 
  left_join(weight, by = "line") %>%
  anyNA()
```

